Gravitational Lens Modelling in a Citizen Science Context

### 2014-10-08 text/ms.tex
edits following comments by SM

The first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y} was developed after
the discovery of the first two gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a massive
galaxy causes a background quasar to be presented as two or four
images.  For the second lens to be discovered (PG1115+080), mass
models scored an early success with the prediction that one of the
lensed images seen would split further into two at higher resolution.
That galaxies must sometimes cause multiple images had long been
expected \citep{1937ApJ....86..217Z}, and it had even been argued that
the phenomenon could help measure cosmological parameters
\citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but apparently nobody
was expecting that lenses would need detailed modelling.  The first
observations, however, immediately stimulated models.  The reason for
that lies in the image separation. Recall that image separations are
of order of the angular Einstein radius
\begin{equation}
\Theta_{\rm E}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $\sim1\rm\,Gpc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses therefore require models of their
mass distributions.

@@

The \sw sims are described in detail in More et al (in prep), but
relevant here is that the sims were of three kinds, as follows.
\begin{enumerate}
\item Lensed quasars: The lens is modelled as a singular isothermal
  ellipsoid (SIE) and a constant external shear whereas the quasar is
  represented with a circular Gaussian source whose size is given by
  the point spread function (PSF) in each imaging band.
\item Galaxy-scale lenses: The lens model is the same as above whereas
  the background galaxy is modelled as an elliptical de~Vaucouleurs.
\item Group-scale lenses: The lens model includes SIE models for the
  central galaxy and the inner group members, plus a circular NFW
  \citep{1996ApJ...462..563N,1997ApJ...490..493N} to represent the
  underlying dark matter distribution and the background galaxy model
  stays the same as galaxy-scale lenses.
\end{enumerate}
The {\tt gravlens} program \citep{2001astro.ph..2340K} was used.
Formulas for the lenses appear in \cite{2001astro.ph..2341K}. The SIE
lenses follow equations (33--35) of that work, with core radius set to
zero.  The NFW lens is in equations (48) and (50), while shear is the
$\gamma$ term in equation (76).

@@

>>>>

The first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y} was developed after
the discovery of the first two gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a massive
galaxy causes a background quasar to appear as two or four
images.  For the second lens to be discovered (PG1115+080), mass
models scored an early success with the prediction that one of the
lensed images seen would split further into two at higher resolution.
That galaxies must sometimes cause multiple images had long been
expected \citep{1937ApJ....86..217Z}, and it had even been argued that
the phenomenon could help measure cosmological parameters
\citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but apparently nobody
was expecting that lenses would need detailed modelling.  The first
observations, however, immediately stimulated models.  The reason for
that lies in the image separation. Recall that image separations are
of order of the angular Einstein radius
\begin{equation}
\Theta_{\rm E}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $\sim1\rm\,Gpc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses therefore require models of their
mass distributions.

@@

The \sw sims are described in detail in More et al (in prep), but
relevant here is that the sims were of three kinds, as follows.
\begin{enumerate}
\item Lensed quasars: The lens is modelled as a singular isothermal
  ellipsoid (SIE) and a constant external shear whereas the quasar is
  represented with a circular Gaussian source whose size is given by
  the point spread function (PSF) in each imaging band.
\item Galaxy-scale lenses: The lens model is the same as above whereas
  the background galaxy is modelled as an elliptical de~Vaucouleurs.
\item Group-scale lenses: The lens model includes SIE models for the
  central galaxy and the inner group members, plus a circular NFW
  \citep{1996ApJ...462..563N,1997ApJ...490..493N} to represent the
  underlying dark matter distribution and the background galaxy model
  stays the same as galaxy-scale lenses.
\end{enumerate}

@@

The {\tt gravlens} program \citep{2001astro.ph..2340K} was used.
Formulas for the lenses appear in \cite{2001astro.ph..2341K}. The SIE
lenses follow equations (33--35) of that work, with core radius set to
zero.  The NFW lens is in equations (48) and (50), while shear is the
$\gamma$ term in equation (76).
